---
title: "Accurate species assignment from 16S amplicon data"
---

--------------------------

The high-resolution afforded by dada2 enables a novel method to classify ribosomal sequence variants (RSVs) with great accuracy: exact matching against a reference database. The `assignSpecies` function implements this functionality, in conjunction with [the species-specific training fasta files](https://www.dropbox.com/sh/mfcivbudmc21cqt/AAB1l-AUM5uKvjrR33ct-cTXa?dl=0) now available for the RDP and Silva databases.

In short, query sequence are compared against all sequences from these databases that had binomial genus-species nomenclature assigned. The genus-species of all exact matches are recorded and returned. Exact matching is not an appropriate method for fuzzy OTUs, as OTUs can include multiple biological variants within their radius. However, is is appropriate for the exact sequences inferred by dada2, and provides accurate assignment at the finest level of taxonomic resolution.

**<span style="color:red">This demonstration requires [version 1.1.2 or later of the dada2 package](https://github.com/benjjneb/dada2):</span>**

```{r load, message=FALSE, warning=FALSE}
library(dada2); packageVersion("dada2")
```

We'll go through a short example to demonstrate the method in action.

*Gzip-ed version of the dada2-formatted species assignment fastas for the [RDP](Training/rdp_species_assignment_14.fa.gz) and [Silva](Training/silva_species_assignment_v123.fa.gz) reference databases are available. Please be aware that the Silva uses a [dual licensing model for academia and commercial users](https://www.arb-silva.de/silva-license-information/).*

&nbsp;

## Species assignment in action

We consider a simple example in which we want to classify a handful of 16S sequences from the v4 region against the RDP training set. Our sequences of interest:
```{r}
seqs <- c(
"ACGGAGGGTCCGAGCGTTAATCGGAATTACTGGGCGTAAAGCGTGCGCAGGCGGTTTGTTAAGCGAGATGTGAAAGCCCCGGGCTCAACCTGGGAATTGCATTTCGAACTGGCGAACTAGAGTCTTGTAGAGGGGGGTAGAATTCCAGGTGTAGCGGTGAAATGCGTAGAGATCTGGAGGA",
"ACAGAGGTCTCAAGCGTTGTTCGGAATCACTGGGCGTAAAGCGTGCGTAGGCTGTTTCGTAAGTCGTGTGTGAAAGGCGCGGGCTCAACCCGCGGACGGCACATGATACTGCGAGACTAGAGTAATGGAGGGGGAACCGGAATTCTCGGTGTAGCAGTGAAATGCGTAGATATCGAGAGGA",
"ACGGAGGGTGCAAGCGTTAATCGGAATCACTGGGCGTAAAGCGCACGTAGGCTGTTATGTAAGTCAGGGGTGAAAGCCCACGGCTCAACCGTGGAACTGCCCTTGATACTGCACGACTCGAATCCGGGAGAGGGTGGCGGAATTCCAGGTGTAGGAGTGAAATCCGTAGATATCTGGAGGA",
"ACGTAGGTCCCGAACGTTGCGCGAATTTACTGGGCGTAAAGGGTCCGTAGGCGGTTTAGCAAGTGGTTGGTGAAATTTCACGGCTCAACCGTGAAACTGCCTTCCAAACTGCTAAACTTGAGGCAGGGAGAGGTCGGCGGAATTCCCGGTGTAGCGGTGAAATGCGTAGATATCGGGAGGA",
"ACGGAGGGGGTTAGCGTTGTTCGGAATTACTGGGCGTAAAGCGTACGTAGGCGGATTGGAAAGTTGGGGGTGAAATCCCAGGGCTCAACCCTGGAACTGCCTCCAAAACTATCAGTCTAGAGTTCGAGAGAGGTGAGTGGAATTCCAAGTGTAGAGGTGAAATTCGTAGATATTTGGAGGA",
"ACGGAGGATCCGAGCGTTATCCGGATTTATTGGGTTTAAAGGGAGCGTAGGTGGACAGTTAAGTCAGTTGTGAAAGTTTGCGGCTCAACCGTAAAATTGCAGTTGATACTGGCTGTCTTGAGTACAGTAGAGGTGGGCGGAATTCGTGGTGTAGCGGTGAAATGCTTAGATATCACGAAGA",
"ACGTAGGGCGCAAGCGTTGTCCGGATTTATTGGGCGTAAAGAGCTCGTAGGCGGCTTGTCGCGTCGACTGTGAAAACCCGTGGCTCAACTGCGGGCTTGCAGTCGATACGGGCAGGCTAGAGTTCGGTAGGGGAGACTGGAATTCCTGGTGTAGCGGTGAAATGCGCAGATATCAGGAGGA",
"ACGGAGGGTGCAAGCGTTAATCGGAATGACTGGGCGTAAAGCGCACGCAGGCGGTCTGTTAAGTTGGATGTGAAATCCCCGGGCTTAACCTGGGAACTGCATTCAAAACTGACAGGCTAGAGTCTCGTAGAGGGGGGTAGAATTCCAGGTGTAGCGGTGAAATGCGTAGAGATCTGGAGGA",
"ACCGACGGCCCGAGTGGTGGCCACTCTTATTGGGCCTAAAGCGTCCGTAGCCGGTCCAGTAAGTCCCTGTTTAAATCCTGCGGCTTAACCGCAGGACTGGCAGGGATACTGCTGGACTTGGGACCGGGAGAGGACAAGGGTACTTCAGGGGTAGCGGTGAAATGTGTTGATCCTTGAAGGA",
"ACGTAGGTCCCGAACGTTGCGCGAAATTACTGGGCGTAAAGGGTCCGTAGGCGGTCTGGTAAGTGGAAGGTGAAAGCCCACGGCTCAACCGTGGAATTGCCTTCCAAACTGCTGGACTTGAGGGCGGAAGAGGTCGGCGGAATTCCCGGTGTAGCGGTGAAATGCGTAGATATCGGGAGGA")
```

We now run the `assignSpecies` function against the RDP training set. *To run the following on your local machine, modify the filename in the second argument to point to where the `rdp_species_assignment_14.fa.gz` file sits on your computer.*

```{r assign-species}
genus.species <- assignSpecies(seqs, "Training/rdp_species_assignment_14.fa.gz")
unname(genus.species)
```

By default the `assignSpecies` method only returns species assignments if there is no ambiguity, i.e. all exact matches were to the same species. However, given that we are generally working with fragments of the 16S gene, it is common that exact matches are made to multiple sequences that are identical over the sequenced region. This is often still useful information, so to have all sequence hits returned the `returnMultiple=TRUE` argument can be passed to the `assignSpecies` function:

```{r assign-species-multi}
unname(assignSpecies(seqs, "Training/rdp_species_assignment_14.fa.gz", allowMultiple=TRUE))
```

Several of these sequences exactly match multiple species. Setting `allowMultiple` to an integer value allows fine-grained control over this behavior. For example, `allowMultiple=3` returns up to 3 different matched species, but if 4 or more are matched it returns 'NA'.

## Integration with `assignTaxonomy`

In many cases it is desirable to link the species annotation in with the broader taxonomic annotation afforded by [`assignTaxonomy`](assign.html). The convenience function `addSpecies` serves this purpose, as it takes as input a taxonomy table, and outputs a table with an added species column. *Note that only those genus-species binomials which are consistent with the genus assigned in the provided taxonomy table are retained in the output.*

```{r add-species}
tt <- assignTaxonomy(seqs, "Training/silva_nr_v123_train_set.fa.gz")
tt.plus <- addSpecies(tt, "Training/silva_species_assignment_v123.fa.gz", verbose=TRUE)
unname(tt.plus)
```

For human microbiome data and V4 sequencing, we typically see a bit under half of the abundant RSVs assigned unambiguously to species-level, and a bit under 2/3rds assigned ambiguously (ie. allowing multiple species assignment). This fraction drops off in less sampled environments and for rare RSVs.

The species assignments of `assignSpecies` are based on **exact sequence matches to cultured representatives**. While what is not in the database can't be matched, the species that are assigned can be considered with higher confidence than the species assignments of inexact comparison methods used for this task elsewhere.