---
title: "DADA2 Installation"
author: "Ben Callahan, Paul McMurdie, Susan Holmes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Installation and Testing}
  \usepackage[utf8]{inputenc}
---


## Installation Instructions

### 0. Install External Dependencies

The following instructions assume that you have installed R 
and have installed 
[Xcode](https://developer.apple.com/xcode/downloads/)
(Mac) or 
[Rtools](http://cran.r-project.org/bin/windows/Rtools/)
(Windows); as well as 
[Bioconductor](http://bioconductor.org/install/).

For Bioconductor,
start a fresh R session, and enter the following.

```{r bioc-install, eval=FALSE}
source("http://bioconductor.org/biocLite.R")
biocLite(suppressUpdates = FALSE)
biocLite("ShortRead", suppressUpdates = FALSE)
```

### 1. Download and unzip.

Download [the zipped package](https://github.com/benjjneb/dada2/archive/master.zip) and unzip it.

### 2. Attempt installation from within R

Start a fresh R session, and enter the following.

```{r install-packages, eval=FALSE}
install.packages("path/to/dada2",
                 repos = NULL,
                 type = "source",
                 dependencies = c("Depends", "Suggests","Imports"))
```

For example, if the dada2 source code directory was in `~/github/dada2`,
then the following would work.

```{r install-github-example, eval=FALSE}
install.packages("~/github/dada2",
                 repos = NULL,
                 type = "source",
                 dependencies = c("Depends", "Suggests","Imports"))
```

Check Package Version

```{r packageVersion}
packageVersion("dada2")
```


### 3. Troubleshoot Dependencies (Optional)

At this point, 
there may be complaints about missing dependencies.
To install missing dependencies on either
[CRAN](http://cran.r-project.org/)
or
[Bioconductor](http://bioconductor.org/install/), 
start a fresh R session, and enter the following.

```{r bioc-install-missing, eval=FALSE}
source("http://bioconductor.org/biocLite.R")
biocLite("missing_package_1")
biocLite("missing_package_2")
# ... and so on
```

### 4. Re-attempt dada2 Installation

Now once again, after dependencies have been installed.

```{r install-packages-rev2, eval=FALSE}
install.packages("path/to/dada2",
                 repos = NULL,
                 type = "source",
                 dependencies = c("Depends", "Suggests","Imports"))
```

You should now be done, if installation was successful.
If not, do panic, and post a message to [the issue tracker](https://github.com/benjjneb/dada2/issues).

### 5. Load Package, Explore Function Documentation

```{r load-dada2, message=FALSE}
library("dada2")
```

Now that the package is loaded,
look at the help files for the main implemented functions.

```{r documentation-example, eval=FALSE, message=FALSE}
help(package="dada2")
?derepFastq
?dada
```

You are ready to use DADA2!
