---
title: "DADA2 Installation"
author: "Benjamin Callahan, Paul McMurdie"
output:
  html_document:
    theme: flatly
---

## 0. Install External Dependencies

The following instructions assume that you have installed R 
and have installed 
[Xcode](https://developer.apple.com/xcode/downloads/)
(Mac) or 
[Rtools](http://cran.r-project.org/bin/windows/Rtools/)
(Windows); as well as 
[Bioconductor](http://bioconductor.org/install/).

For Bioconductor,
start a fresh R session, and enter the following.

```{r bioc-install, eval=FALSE}
source("http://bioconductor.org/biocLite.R")
biocLite(suppressUpdates = FALSE)
biocLite("ShortRead", suppressUpdates = FALSE)
```

## 1. Install using devtools

The easiest method is to use the devtools package to install from github:

```{r devtools-install, eval=FALSE}
install.packages("devtools")
library("devtools")
devtools::install_github("benjjneb/dada2")
```

If that worked you're done (see step 4)! If it failed on missing dependencies, goto step 2.

## 1a. Install manually from source

Alternatively, you can install from the source by hand. Download [the zipped package](https://github.com/benjjneb/dada2/archive/master.zip) and unzip it.

Start a fresh R session, and enter the following.

```{r install-packages, eval=FALSE}
install.packages("path/to/dada2",
                 repos = NULL,
                 type = "source",
                 dependencies = c("Depends", "Suggests","Imports"))
```

For example, if the dada2 source code directory was in `~/github/dada2`,
then the following would work.

```{r install-github-example, eval=FALSE}
install.packages("~/github/dada2",
                 repos = NULL,
                 type = "source",
                 dependencies = c("Depends", "Suggests","Imports"))
```

Check Package Version

```{r packageVersion}
packageVersion("dada2")
```

## 2. Troubleshoot Dependencies (Optional)

At this point, 
there may be complaints about missing dependencies.
To install missing dependencies on either
[CRAN](http://cran.r-project.org/)
or
[Bioconductor](http://bioconductor.org/install/), 
start a fresh R session, and enter the following.

```{r bioc-install-missing, eval=FALSE}
source("http://bioconductor.org/biocLite.R")
biocLite("missing_package_1")
biocLite("missing_package_2")
# ... and so on
```

## 3. Re-attempt dada2 Installation

Install again, after dependencies have been installed.

Devtools method:
```{r devtools-install-rev2, eval=FALSE}
library("devtools")
devtools::install_github("benjjneb/dada2")
```

Manual install from source method:
```{r install-packages-rev2, eval=FALSE}
install.packages("path/to/dada2",
                 repos = NULL,
                 type = "source",
                 dependencies = c("Depends", "Suggests","Imports"))
```

You should now be done, if installation was successful.
If not, do not panic, and post a message to [the issue tracker](https://github.com/benjjneb/dada2/issues).

## 4. Load Package, Explore Function Documentation

```{r load-dada2, message=FALSE}
library("dada2")
```

Now that the package is loaded,
look at the help files for the main implemented functions.

```{r documentation-example, eval=FALSE, message=FALSE}
help(package="dada2")
?derepFastq
?dada
```

You are ready to use DADA2! Consider starting by taking a look at our [tutorial on paired-end Miseq data](tutorial.html).
